dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format

### 2023-12-15 ms.Qmd
Rearrange examples

The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DwC) standard [@Wieczorek2012] for biodiversity data.
`dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
Furthermore, it conducts automated checks to validate taxonomic datasets, thereby alerting database maintainers to problems as soon as possible.
`dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.
`dwctaxon` has passed [code review](https://github.com/ropensci/software-review/issues/574) at [ROpenSci](https://ropensci.org/).
Complete documentation is available at <https://docs.ropensci.org/dwctaxon/>.

@@

## Example usage

First, we create a small fictional dataset in the DwC format (DwC possibly includes many more columns, but we only use a subset here) for demonstration using the `tibble` package:

```{r}
#| label: load-pkg-data
#| message: false
library(tibble)
library(dwctaxon)

example_data <- tribble(
  ~taxonID, ~acceptedNameUsageID, ~taxonomicStatus, ~taxonRank, ~scientificName,
  "id1", NA, "accepted", "species", "Fooius barus",
  "id2", "id1", "synonym", "species", "Fooius wizum",
  "id3", NA, "accepted", "species", "Bazius quxus",
  "id4", NA, "accepted", "species", "Bazius buzbar"
)
```

```{r}
#| label: dct-options
#| echo: false

# Store existing settings
old_settings <- dct_options()

# turn off time stamp
dct_options(stamp_modified = FALSE)
```

`dct_add_row()` adds one or more rows.
If the user does not provide a value for `taxonID`, one is generated from the hash of the scientific name, such that it should be unique (as long as the scientific name is unique).

```{r add-row}
example_data |>
  dct_add_row(
    scientificName = "Bazius qixis",
    taxonomicStatus = "accepted",
    taxonRank = "species"
  )
```

`dct_modify_row()` modifies a row, automatically re-mapping synonyms if needed.
Here, we change *Bazius buzbar* to be a synonym of *Bazius qixis* (note that `acceptedNameUsageID` is automatically populated):

```{r modify-row}
example_data |>
  dct_modify_row(
    scientificName = "Bazius buzbar",
    taxonomicStatus = "synonym",
    acceptedNameUsage = "Bazius quxus"
  )
```

`dct_fill_col()` fills in values for columns that have term-termID pairs.
Here, we demonstrate filling in `acceptedNameUsage` (the resulting values do not appear in the manuscript due to width restrictions, but can be seen by running the code interactively).

```{r fill-col}
example_data |>
  dct_fill_col(
    fill_to = "acceptedNameUsage",
    fill_from = "scientificName",
    match_to = "taxonID",
    match_from = "acceptedNameUsageID"
  )
```

The `example_data` dataset is already well-formatted, so it will pass validation:

```{r validate-pass}
# Default behavior is to return the original dataset if checks pass
# For this example, return TRUE instead
dct_validate(example_data, on_success = "logical")
```

For demonstration purposes, we create a dataset with duplicated and erroneous entries:

```{r make-dirty-filmies}
# Duplicate some data
bad_data <- rbind(head(example_data), head(example_data, 2))
# Replace some values of `acceptedNameUsageID` with random letters
bad_data$acceptedNameUsageID[sample(seq_along(bad_data), 3)] <- sample(letters, 3)
```

By default, `dct_validate()` will stop with an error on the first check that fails:

```{r validate-error, error = TRUE}
dct_validate(bad_data)
```

But it may be useful to get an overview of all the checks that failed.
This can be done by setting `on_fail` to `"summary"`:

```{r validate-summary-show, eval = FALSE, echo = TRUE}
dct_validate(bad_data, on_fail = "summary")
```

```{r validate-summary-print, echo = FALSE, message = FALSE}
dct_validate(bad_data, on_fail = "summary") |>
  dplyr::mutate(error = stringr::str_trunc(error, 40, "right"))
```

### Piping

All the functions in dwctaxon take a dataframe as their first argument and return a dataframe by default, so they are "pipe-friendly" and can be chained together:

```{r pipe}
example_data |>
  dct_modify_row(
    taxonID = "id2",
    taxonomicStatus = "accepted"
  ) |>
  dct_add_row(
    scientificName = "Bazius qixis",
    taxonomicStatus = "accepted"
  ) |>
  dct_validate()
```

It's often a good idea to include `dct_validate()` at the end of a chain to make sure the modified taxonomic database is still correctly formatted.

### Real-world example

The small examples shown above may not reflect the performance of `dwctaxon` when applied to larger "real-world" datasets.
Therefore, in this section we demonstrate usage of `dwctaxon` with the GBIF backbone taxonomy, which is to our knowledge the largest single database in DwC format currently available online, with `r number(gbif_backbone_size, big.mark = ",")` rows (names) in the most recent release (2023-08-28) as of writing.

The taxonomy is loaded into R with `readr::read_tsv()`, then validated with `dct_validate()`.
```{r}
#| label: demo-gbif
#| eval: false

@@

```{r}
#| label: demo-gbif-hide
#| echo: false
warning(
  targets::tar_meta(fields = warnings, complete_only = TRUE)$warnings %>%
  str_replace_all("\\.\\. ", ". ") %>%
  str_split("\\. ") %>%
  magrittr::extract2(1) %>%
  paste0(".") %>%
  map_chr(~str_wrap(., width = 70)) %>%
  paste(collapse = "\n")
)
```

This reveals a large number of potential issues with the data, as shown by the warnings.
Furthermore, since all potential problems with the data have been saved to a dataframe by setting `on_fail = "summary"`, we can dig into the output for a more detailed view.
Here, we demonstrate this using the tidyverse:

```{r}
#| label: error-count-show
#| message: false
library(tidyverse)
gbif_check %>%
  count(error)
```

```{r}
#| label: error-count-hide
#| echo: false
gbif_check_count <-
  gbif_check %>%
  count(error)

n_err <- function(match_text) {
  n <- gbif_check_count %>%
    filter(str_detect(error, match_text)) %>%
    pull(n) %>%
    sum()
  stopifnot(length(n) == 1)
  if (n > 10) {
    return(scales::number(n, big.mark = ","))
  } else {
    return(as.character(english::english(n)))
  }
}
```

@@

Another real-world example using the Database of Vascular Plants of Canada [@Desmet2013] is available on the `dwctaxon` [documentation website](https://docs.ropensci.org/dwctaxon/articles/real-data.html).

## Caveats

```{r}
#| label: caveats
#| echo: false

times <- targets::tar_meta(fields = seconds)
check_time <-
  times %>%
  filter(name == "gbif_check") %>%
  pull(seconds) %>%
  round(1) %>%
  paste("seconds")
```

One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.
However, our demonstration with GBIF data shows that `dwctaxon` functions can be carried out in a reasonable amount of time (in this case, `r check_time` on an iMac) on a ca. 2 Gb database that includes a large portion of all scientific names currently in use.
Therefore, there are likely to be few situations where `dwctaxon` cannot be used due to database size.

@@

```{r}
#| label: dct-options-reset
#| echo: false

# Reset options to previous
do.call(dct_options, old_settings)
```

>>>>

The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DwC) standard [@Wieczorek2012] for biodiversity data.
`dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
Furthermore, it conducts automated checks to validate taxonomic datasets, thereby alerting database maintainers to problems as soon as possible.
`dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.
`dwctaxon` has passed [code review](https://github.com/ropensci/software-review/issues/574) at [ROpenSci](https://ropensci.org/).
Complete documentation is available at <https://docs.ropensci.org/dwctaxon> and source code is available at <https://github.com/ropensci/dwctaxon>.

@@

## Examples and documentation

Examples of code usage, an overview of the DwC taxon data format, and complete description of all functions are available on the `dwctaxon` website, <https://docs.ropensci.org/dwctaxon>.
The examples include [editing](https://docs.ropensci.org/dwctaxon/articles/editing.html) and [validating](https://docs.ropensci.org/dwctaxon/articles/validation.html) a small dataset that comes with `dwctaxon` as well as a [real-world example](https://docs.ropensci.org/dwctaxon/articles/real-data.html) using the Database of Vascular Plants of Canada [@Desmet2013].

## Benchmark on real data

One potential limitation of data validation that relies on in-memory data loaded into R (via `load.csv()` etc.) is that extremely large databases could exceed R's memory capacity.

To test this, we ran `dct_validate()` on the GBIF backbone taxonomy, which is to our knowledge the largest single database in DwC format currently available online with `r number(gbif_backbone_size, big.mark = ",")` rows (names) in the most recent release (2023-08-28) as of writing.
The test was carried out on an Apple iMac (2019, 16 GB RAM) with R v`r getRversion()` and `dwctaxon` v`r packageVersion("dwctaxon")`.
The code and results are reproduced below:

```{r}
#| label: demo-gbif
#| eval: false

@@

```{r}
#| label: demo-gbif-hide
#| echo: false
# Reproduce warnings picked up during analysis
warning(
  targets::tar_meta(fields = warnings, complete_only = TRUE)$warnings %>%
  str_replace_all("\\.\\. ", ". ") %>%
  str_split("\\. ") %>%
  magrittr::extract2(1) %>%
  paste0(".") %>%
  map_chr(~str_wrap(., width = 70)) %>%
  paste(collapse = "\n")
)
```

```{r}
#| label: demo-gbif-table-show
#| eval: FALSE
# Count types of errors detected
table(gbif_check$error)
```

```{r}
#| label: demo-gbif-table-hide
#| echo: false
#| eval: TRUE
# Wrap error table names to fit PDF width
error_table <- table(gbif_check$error)
error_table_names <-
  names(error_table) %>%
  str_wrap(width = 70)
for (i in seq_along(error_table)) {
  cat(error_table_names[[i]])
  cat("\ \ \n")
  cat(error_table[[i]])
  cat("\ \ \n")
}
```

```{r}
#| label: error-count-hide
#| echo: false

# Get running times
times <- targets::tar_meta(fields = seconds)
check_time <-
  times %>%
  filter(name == "gbif_check") %>%
  pull(seconds) %>%
  round(1) %>%
  paste("seconds")

gbif_check_count <-
  gbif_check %>%
  count(error)

# Helper function to print number of errors
n_err <- function(match_text) {
  n <- gbif_check_count %>%
    filter(str_detect(error, match_text)) %>%
    pull(n) %>%
    sum()
  stopifnot(length(n) == 1)
  if (n > 10) {
    return(scales::number(n, big.mark = ","))
  } else {
    return(as.character(english::english(n)))
  }
}

```

@@

The time to run `dct_validate()` even on this relatively large database was quite short: `r check_time`.
Therefore, there are likely to be few situations where `dwctaxon` cannot be used due to database size.

@@

## Reproducibility

Code to generate this manuscript is available at <https://github.com/joelnitta/dwctaxon_ms>.
